---
title: "HW1"
author: "Nami Niyakan"
date: "9/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## scBALF Covid-19 data analysis with Monocle

Setting up monocle model by using the normal distribution. (exexpressionFamily=uninormal(), default is the negativeBinomial)




```{r cars}
knitr::opts_chunk$set(cache = TRUE)
library(monocle)
library(dplyr)
data<-read.csv(file = "/Users/Nami/Documents/hackathon/covid-selected-data.csv")
data<-as.matrix(data)
rownames(data)<-data[,1]
data<-data[,-1]
n<-rownames(data)
data<-apply(data,2,as.numeric)
rownames(data)<-n
data<-t(data)
dim(data)

true_lables<-read.csv(file="/Users/Nami/Documents/hackathon/covid-selected-data-labels.csv")
true_lables<-as.matrix(true_lables)
rownames(true_lables)<-true_lables[,1]
true_lables<-true_lables[,-1]
n<-rownames(data)
rownames(data)<-n
true_lables<-as.matrix(true_lables)
colnames(true_lables)<-c("cell_type")
true_lables<-as.data.frame(true_lables)
summary(as.factor(true_lables$cell_type))

true_lables<-true_lables %>% mutate(Nor_vs_rest = case_when(
  cell_type == c("Normal") ~ "Normal",
  cell_type != c("Normal") ~ "Rest"))

true_lables<-true_lables %>% mutate(Sev_vs_rest = case_when(
  cell_type == c("Severe") ~ "Severe",
  cell_type != c("Severe") ~ "Rest"))

features<-matrix(0,nrow(data),1)
colnames(features)<-c("gene_short_name")
rownames(features)<-rownames(data)
features[,1]<-rownames(data)
features<-as.data.frame(features)

```

```{r cars1}
pd <- new("AnnotatedDataFrame", data = true_lables)
fd <- new("AnnotatedDataFrame", data = features)
covid <- newCellDataSet(as.matrix(data),
                       phenoData = pd, featureData = fd,expressionFamily=uninormal())

```




## Dimension Reduction using tSNE and cell clustering (densityPeak)


```{r pressure}
knitr::opts_chunk$set(cache = TRUE)
covid <- reduceDimension(covid, max_components = 2, num_dim = 50,
                         norm_method = 'none',
                        reduction_method = 'tSNE',
                        residualModelFormulaStr = "~cell_type",
                        verbose = F)
covid <- clusterCells(covid, num_clusters = 4)
plot_cell_clusters(covid, 1, 2, color = "cell_type")
plot_cell_clusters(covid, 1, 2, color = "Cluster")
plot_cell_clusters(covid, 1, 2, color = "Cluster") +
  facet_wrap(~cell_type)
```

```{r cars2, echo=FALSE}
pd <- new("AnnotatedDataFrame", data = true_lables)
fd <- new("AnnotatedDataFrame", data = features)
covid <- newCellDataSet(as.matrix(data),
                       phenoData = pd, featureData = fd,expressionFamily=uninormal())
```



## Dimension Reduction using tSNE and cell clustering(densityPeak)

This time used 200 principal components instead of 50 in the initial PCA to see whether results change or not. Also removed adding pseudo expression and relative expression options.

```{r pressure1}
covid <- reduceDimension(covid, max_components = 2, num_dim = 200,
                         norm_method = 'none',
                        reduction_method = 'tSNE',
                        pseudo_expr = 0,
                        relative_expr = FALSE,
                        residualModelFormulaStr = "~cell_type",
                        verbose = F)
covid <- clusterCells(covid, num_clusters = 4)
plot_cell_clusters(covid, 1, 2, color = "cell_type")
plot_cell_clusters(covid, 1, 2, color = "Cluster")
plot_cell_clusters(covid, 1, 2, color = "Cluster") +
  facet_wrap(~cell_type)

```


```{r cars3,echo=FALSE}
pd <- new("AnnotatedDataFrame", data = true_lables)
fd <- new("AnnotatedDataFrame", data = features)
covid <- newCellDataSet(as.matrix(data),
                       phenoData = pd, featureData = fd,expressionFamily=uninormal())
```

## Dimension Reduction using tSNE and cell clustering(Louvain with K=1000 in knn)

```{r pressure2}
covid <- reduceDimension(covid, max_components = 2, num_dim = 50,
                         norm_method = 'none',
                        reduction_method = 'tSNE',
                        pseudo_expr = 0,
                        residualModelFormulaStr = "~cell_type",
                        relative_expr = FALSE,
                        verbose = F)
covid <- clusterCells(covid, method = 'louvain',k=1000)
plot_cell_clusters(covid, 1, 2, color = "cell_type")
plot_cell_clusters(covid, 1, 2, color = "Cluster")
plot_cell_clusters(covid, 1, 2, color = "Cluster") +
  facet_wrap(~cell_type)
```
```{r cars4,echo=FALSE}
pd <- new("AnnotatedDataFrame", data = true_lables)
fd <- new("AnnotatedDataFrame", data = features)
covid <- newCellDataSet(as.matrix(data),
                       phenoData = pd, featureData = fd,expressionFamily=uninormal())
```


## Differential Expression Analysis between Normal cells and the rest (Mild + Severe)
```{r pressure3}
marker_genes <- row.names(subset(fData(covid),
                                 gene_short_name %in% gene_short_name))
diff_test_res_Nor <- differentialGeneTest(covid[marker_genes,],
                                      fullModelFormulaStr = "~Nor_vs_rest")
Ordered_diff_test_res_Nor<- diff_test_res_Nor[order(diff_test_res_Nor$qval),]
head(Ordered_diff_test_res_Nor,20)
```

## Differential Expression Analysis between Severe cells and the rest (Mild + Normal)
```{r pressure4}

diff_test_res_Sev <- differentialGeneTest(covid[marker_genes,],
                                      fullModelFormulaStr = "~Sev_vs_rest")
Ordered_diff_test_res_Sev<- diff_test_res_Sev[order(diff_test_res_Sev$qval),]
head(Ordered_diff_test_res_Sev,20)
```

Intersection of top 100 DE genes between two analysis:

```{r pressure5}
intersect(Ordered_diff_test_res_Nor$gene_short_name[1:100],Ordered_diff_test_res_Sev$gene_short_name[1:100])
```

